\documentclass{llncs}
\input{"../preamble.tex"}
\begin{document}

\section{Related Work}

As the first pattern mining approach using the MDL principle, Krimp \cite{krimp} was one of the main sources of inspiration for this paper. Many papers on pattern-based modelling using MDL have appeared since, both improving search, e.g., Slim \cite{slim}, and extensions to other problems, e.g., Classy \cite{classy} for rule-based classification.

The problem closest to ours is probably that of geometric tiling, as introduced by Gionis et al.\ \cite{gionis2004tiles} and later also combined with the MDL principle by Tatti and Vreeken \cite{tatti2012stijl}. Geometric tiling, however, is limited to Boolean data and rectangularly shaped patterns (tiles); we strongly relax both these limitations (but as of yet do not support patterns based on densities or noisy occurrences).

Campana et al. \cite{campana2010compression} also use matrix-like data (textures) in a compression-based similarity measure. Their method, however, has less value for \emph{explanatory} analysis as it relies on generic compression algorithms that are essentially a black box.

Geometric pattern mining is different from graph mining, although the concept of a matrix can be redefined as a grid-like graph where each node has a fixed degree. This is the approach taken by Deville et al. \cite{deville2016grima}, solving a problem similar to ours but using an approach akin to bag-of-words instead of the MDL principle.

\end{document}
